---
title: "Determining a Fair Apartment Price"
subtitle: "Stat 410 Final Project"
author: "Jack Dzialo, Devin Abraham, Tom Vincent"
format:
  revealjs:
    theme: ./_extensions/grantmcdermott/clean/clean.scss
---

## Purpose

### Why choose our dataset?

TODO: CI of sample house around Rice

::: {.incremental}
- Aimed to address an [actual]{.bg style="--col: #D0E1F9"} problem that we face
- Currently looking for housing next year
:::

::: {.fragment .r-fit-text}
[Challenge]{.bg style="--col: #D0E1F9"}: How to determine if price is fair?
:::

## Proposal

::: {.incremental}
- Chose to use [Forward Stepwise Search]{.bg style="--col: #D0E1F9"}, using F tests
  - Tested interaction terms once individual predictors were determined
- Read up upon different ways to determine model accuracy
- Use confidence intervals generated by the model to see if posted price for an apartment is fair
:::


::: {.fragment}
::: {.callout-tip}
## Definitions
1. **AIC - **
2. **AICc - **
3. **BIC - ** hello
4. **PRESS - ** Predicted Residual Error Sum of Squares; predicts how accurate a model is at predicting data it hasn't seen before
:::
:::

## Data Collection

::: {.incremental}
- Dataset of apartments for rent, sourced from UCI dataset repository
- [100,000]{.bg style="--col: #D0E1F9"} rows, each representing a different apartment
- [22]{.bg style="--col: #D0E1F9"} columns, each representing a descriptor for an apartment listing
  - Square feet, apartment type, number of bathrooms, has photos
:::
::: {.fragment}
::: {.columns}

::: {.column width="50%"}
```{r fig.width=8, fig.height=6}
data <- read.csv2("apartments.csv")
hist(as.numeric(data$price),
     freq = FALSE,
     breaks = "scott",
     main = "Histogram of Price",
     xlab = "Price (dollars)",
     col = "lightblue",
     xlim = c(0, 10000))
```
:::
::: {.column width="50%"}
::: {.centered}
```{r}
data <- read.csv2("apartments.csv")
data$price <- as.numeric(data$price)
table <- data.frame(
  Statistic = c("Mean", "Max", "Min", "Median"),
  Value = c(round(mean(data$price, na.rm = TRUE)),
            round(max(data$price, na.rm = TRUE)),
            round(min(data$price, na.rm = TRUE)),
            round(median(data$price, na.rm = TRUE))
))
knitr::kable(table)
```
:::
:::
:::
:::

# Data Analysis - Model Building
